### 2017-03-21 main.tex
Ben's final(ish) round of edits. Still need to provide DOIs for code and data repositories, and update diffs.

If precise moments in song can reliably be detected quickly enough to
trigger other apparatus during singing, then this consistency of
behaviour allows a variety of experiments.  A common area of study
with song-triggered experiments is the anterior forebrain pathway
(AFP), a homologue of mammalian basal ganglia consisting of a few
distinct brain areas concerned with the learning and production of
song.  For example, stimulation of the lateral magnocellular nucleus
of the anterior nidopallium (LMAN)---the output nucleus of the
AFP---at precisely timed moments during song showed that this area
controls specific variables in song output \cite{Kao2005}.
Song-synchronised stimulation of LMAN and the high vocal centre (HVC)
in one hemisphere or the other showed that control of song rapidly
switches between hemispheres \cite{Wang2008interhemispheric}.
Feedback experiments have shown that Field L and the caudolateral
mesopallium may hold a representation of song against which auditory
signals are compared \cite{Keller2009}.  Renditions of a syllable that
were disrupted by white noise when they were slightly above (or below)
the syllable's average pitch showed that the apparently random natural
variability in songbird motor output is used to drive change in the
song \cite{Tumer2007pitchshift}. Furthermore, AFP produces a
corrective signal to bias song away from those disruptions
\cite{Andalman2009}. The song change is isolated to within roughly 10
milliseconds (ms) of the stimulus, and the shape of the learned
response can be predicted by a simple mechanism
\cite{Charlesworth2011learning}. AFP transfers the error signal to the
robust nucleus of the arcopallium (RA) using NMDA-receptor--mediated
glutamatergic transmission \cite{Warren2011}. The course of song
recovery after applying such a pitch-shift paradigm showed that the
caudal medial nidopallium is implicated in memorising or recalling a
recent song target, but in neither auditory processing nor directed
motor learning \cite{Canopoli2014}.

@@

Despite the power and versatility of vocal feedback experiments, there
is no standard syllable detector.  Desiderata for such a detector include:
\begin{description}
\item[Accuracy:] How often does the system produce false positives or false negatives?
\item[Latency:] The average delay between the target syllable being sung and the detection.
\item[Jitter:] The amount that latency changes from instance to instance of song.  Our measure of jitter is the standard deviation $\sigma$ of latency.
\item[Versatility:] Is detection possible at ``difficult'' syllables?
\item[Ease of use:] How automated is the process of programming a detector?
\item[Cost:] What are the hardware and software requirements?
\end{description}

A variety of syllable-triggering systems have been used, but few have
been documented or characterised in detail.  In 1999,
detection was achieved by a group of IIR filters with hand-tuned logical
operators \cite{Leonardo1999}.  The system had a latency of 50 or 100 ms,
and jitter or accuracy were not reported.  As access to
computational resources has improved, approaches have changed: in
2009, hand-tuned filters were implemented on a Tucker-Davis Technologies digital signal processor, bringing latency down to around 4 ms \cite{Andalman2009}. But as with other filter-bank techniques, it
is not strictly a syllable detector but rather a pitch and timbre
detector---it cannot identify a frequency sweep, or distinguish a
short chirp from a long one---and thus requires careful selection of
target syllables.  Furthermore, the method is neither inexpensive nor,
based on our experience with a similar technique, accurate.
2009 also saw the application of a neural network to a spectral image of
song \cite{Keller2009}.  They reported a jitter of 4.3 ms, but further implementation and
performance details are not available.  In 2011, stable portions of syllables were matched to spectral templates in 8-ms
segments \cite{Warren2011}.  This detector achieved a jitter of 4.5 ms, and false-negative and
false-positive rates of up to 2\% and 4\% respectively.  Hardware
requirements and ease of use were not reported.  In 2013, spectral
images of template syllables were compared to song using a correlation coefficient \cite{Skocik2013}.  With a
fast desktop (Intel i7 six-core) running Linux and equipped with a
National Instruments data acquisition card, it boasts a hardware-only
(not accounting for the time taken to compute a match with a syllable)
latency and jitter of just a few microseconds, and the detection
computation they use should not much increase that.  They reported
false-negative rates around 4\% and 7\% for zebra finches and
Bengalese finches respectively, measured on a small dataset.  In much
other work, a syllable detector is alluded to, but not described.

@@

\begin{enumerate}
\item Record and align a corpus of training songs.  The technique has been published in \cite{Poole2012}. 200 songs generally yield acceptable results, but here we standardise on 1000-song training sets.
\item Choose one or more instants in the song that should create a trigger event, and train a
  neural network to recognise them. This step is carried out offline.  While any neural network software would produce similar results, we used MATLAB 2015b's neural network toolbox.
\item Once trained and saved, the neural network is used by a realtime
  detection programme that listens to an audio signal and indicates detection of the target
  syllables via a TTL pulse.  We present three detection implementations, in MATLAB, LabVIEW, and Swift, that trade off hardware requirements, ease of maintenance, and performance.
\end{enumerate}
This method makes the following contributions:
\begin{itemize}
\item Fast: sub-millisecond latencies, with jitter around 2 ms.
\item Accurate: false negative rates under 1\% and false positive rates under 0.005\% for a variety of syllables.
\item State-of-the-art performance with default parameters.
\item Requires almost no programming experience.
\item Runs on inexpensive hardware.
\item Described in detail here, with reference implementations provided and benchmarked.
\end{itemize}

@@

We begin with a few hundred recordings of a given bird's song, as well
as calls, cage noise, and other non-song audio data.  Male zebra finch
song is most highly stereotyped when a female is present (``directed
song'') and slightly more variable otherwise (``undirected''); we
train and test on undirected song since this is more commonly studied,
as well as more challenging.  Recordings were made inside a
sound-isolating chamber in which was mounted a recording microphone
(Audio-Technica AT803), using methods similar to those described in \cite{Tchernichovski2011SAP}, Chapter 2.  The songs were
time-aligned as described in \cite{Poole2012}.

@@

Six examples of chosen target moments in the song, with recognition
regions $F = 1--8$ kHz and $T=30$ ms, are shown in \fig{fig:song}.

@@

With inputs well outside the space on which a neural network has been
trained, its outputs will be essentially random. In order to reduce
the false positive rate it is necessary to provide negative training
examples that include silence, cage noise, wing flapping, non-song
vocalisations, and perhaps songs from other birds.  Although it will
depend on the makeup of the non-song data, we have found that training
with as low as a 1:1 ratio of non-song to song---or roughly 10
minutes of non-song audio---yields excellent results on most birds.
These data are also included in our accuracy tests.

@@

During runtime, the network may produce above-threshold responses to nearby
frames.  Thus, after the first response, subsequent responses are suppressed
for 100 ms.  However, for the accuracy figures presented here, we used
the un-de-bounced network response.

@@

This implementation requires special software and hardware: LabVIEW
from National Instruments---we used 2014 service pack 1---and a data acquisition card---we use the
National Instruments PCI-6251 card on a PC with an Intel Core i5-4590
processor at 3.7GHz (a relatively low-end machine) with 32 gigabytes
of RAM, running Microsoft Windows 8.1 Pro and Windows 10.

@@

We define the accuracy of the network based on its classification
performance per frame. In order to avoid the apparent problem of every
non-detected non-syllable counting as a true negative, we also tried
defining accuracy on a per-song basis, such that a song-length sample without the
target syllable counted as a single true negative.  Computing the
optimal output thresholds on a per-frame rather than a per-song basis resulted in higher
thresholds and thus a slight reduction in both the false-positive and true-positive rates (reversible by increasing $C_n$), while also providing a valid
definition of false-positive rate for data streams that had not been
segmented into song-sized chunks and thus allowing easy interpretation
of false-positive rates from calls, cage noise, etc.

@@

\fig{fig:song} shows a typical song. The image is produced by
averaging the aligned spectrograms for the bird lny64.  Six target
trigger points are shown for this song, at 150, 200, 250, 300, 350,
and 400 milliseconds after the beginning of the aligned samples, as
indicated by the red lines. In discussion, we will refer to these six
as $t^*_1\ldots t^*_6$ respectively.  For clarity, the rectangles show
a shorter recognition window than our standard: here 30 ms, but still
spanning 1--8 kHz, for each target time in the song.\marginpar{Is this
  OK?  Should I update to a 50-ms window?  They'll overlap.}

@@

\begin{table}
  % replot_accuracies_concatanated('tex', true) for confusion_log_perf.txt (1.5 ms).
  \begin{tabular}{lc|cc}
    Bird & Target time (ms) & \% True positives & \% False positives \\ \hline
lny64 & 	150 $[t^*_1]$ & 	99.66 & 	0.0026 \\ % /Volumes/Data/song/lny64/confusion_log_perf.txt
      & 	200 $[t^*_2]$ & 	99.75 & 	0.00052 \\ %                                               
      & 	250 $[t^*_3]$ & 	99.79 & 	0.00051 \\ %                                               
      & 	300 $[t^*_4]$ & 	99.70 & 	0.00077 \\ %                                               
      & 	350 $[t^*_5]$ & 	99.73 & 	0.00022 \\ %                                               
      & 	400 $[t^*_6]$ & 	99.52 & 	0.0010 \\ %                                                
lny42 & 	380 & 	99.09 & 	0.0011 \\ % /Volumes/Data/song/lny42/confusion_log_perf.txt          
lny46 & 	140 & 	99.53 & 	0.0041 \\ % /Volumes/Data/song/lny46/confusion_log_perf.txt          
lny4rb & 	250 & 	99.36 & 	0.0025 \\ % /Volumes/Data/song/lny4rb/confusion_log_perf.txt        
lr12 & 	180 & 	99.52 & 	0.0033 \\ % /Volumes/Data/song/lr12/all/confusion_log_perf.txt        
lr13 & 	300 & 	99.83 & 	0.0013 \\ % /Volumes/Data/song/lr13/all/confusion_log_perf.txt        
lr28 & 	140 & 	99.73 & 	0.0028 \\ % /Volumes/Data/song/lr28/all/confusion_log_perf.txt        
lr77 & 	320 & 	99.58 & 	0.0028  % /Volumes/Data/song/lr77/all/confusion_log_perf.txt          
  \end{tabular}
  \caption{Mean values for the detection accuracies shown in \fig{fig:accuracies}.}
  \label{table:accuracies}
\end{table}

@@

Changing the FFT interval to 2 ms significantly reduces the
computational requirement. We found that some syllables were slightly
easier to detect and some more difficult, with no significant
difference on average.

@@

  \begin{description}
  \item[Installation:]\hfill
    \begin{itemize}
    \item Clone the Github repository onto your computer.
    \item You also need the following packages from MATLAB File Exchange somewhere in your MATLAB path:
      \begin{description}
      \item[Significant Figures] by Teck Por Lam.  File ID \#10669.
      \item[Generate maximally perceptually-distinct colors] by Tim Holy.  File ID \#29702.
      \end{description}
    \end{itemize}
  \item[Usage:]\hfill
    \begin{itemize}
      \item The main programme is {\tt learn\_detector.m}.
    \item The top section of {\tt learn\_detector.m} contains global
      configuration variables.  The important ones are the data
      directory and the bird name: as supplied, the software looks for
      {\tt \textless data directory\textgreater/\textless bird
        name\textgreater/song.mat}.  Modify as necessary.
    \item The format for the data file {\tt \textless data
      directory\textgreater/\textless bird name\textgreater/song.mat}
      is specified in {\tt syllable-detector-learn/README.md}: a
      MATLAB file containing aligned song, nonsong (calls, cage noise,
      etc), and the audio sampling frequency.
    \item The configuration file {\tt \textless data
      directory\textgreater/\textless bird name\textgreater/params.m}
      is also described in {\tt README.md}.  If the file is empty, the
      programme will display an image showing the average of the song
      spectrograms (frequency vs.~time), tell the user how to define
      syllables of interest, and exit.  If {\tt times\_of\_interest}
      is defined, the programme will train the neural network.
      Optionally, {\tt params.m} can also override any of the
      configuration parameters at the top of {\tt learn\_detector.m}.
    \item The software will produce a neural network file, {\tt
      detector\_bird\_times\_\dots.mat} (the ellipses signify other
      information encoded in the filename).  It will also produce the
      audio test file described in Methods.
    \end{itemize}
  \end{description}

>>>>

If precise moments in song can reliably be detected quickly enough to
trigger other apparatus during singing, then this consistency of
behaviour allows a variety of experiments.  A common area of study
with song-triggered experiments is the anterior forebrain pathway
(AFP), a homologue of mammalian basal ganglia consisting of a few
distinct brain areas concerned with the learning and production of
song.  For example, stimulation of the lateral magnocellular nucleus
of the anterior nidopallium (LMAN)---the output nucleus of the
AFP---at precisely timed moments during song showed that this area
controls specific variables in song output \cite{Kao2005}.
Song-synchronised stimulation of LMAN and the high vocal centre (HVC)
in one hemisphere or the other showed that control of song rapidly
switches between hemispheres \cite{Wang2008interhemispheric}.
Feedback experiments have shown that Field L and the caudolateral
mesopallium may hold a representation of song against which auditory
signals are compared \cite{Keller2009}.  The disruption by white noise
of renditions of a syllable that were slightly above (or below) the
syllable's average pitch showed that the apparently random natural
variability in songbird motor output is used to drive change in the
song \cite{Tumer2007pitchshift}, and AFP produces a
corrective signal to bias song away from those disruptions
\cite{Andalman2009}. The song change is isolated to within roughly 10
milliseconds (ms) of the stimulus, and the shape of the learned
response can be predicted by a simple mechanism
\cite{Charlesworth2011learning}. AFP transfers the error signal to the
robust nucleus of the arcopallium (RA) using NMDA-receptor--mediated
glutamatergic transmission \cite{Warren2011}. The course of song
recovery after applying such a pitch-shift paradigm showed that the
caudal medial nidopallium is implicated in memorising or recalling a
recent song target, but in neither auditory processing nor directed
motor learning \cite{Canopoli2014}.

@@

Despite the power and versatility of vocal feedback experiments, there
is no standard syllable detector.  Desiderata for such a detector include:
\begin{description}
\item[Accuracy:] How often does the system produce false positives or false negatives?
\item[Latency:] The average delay between the target syllable being sung and the detection.
\item[Jitter:] The amount that latency changes from instance to instance of song.  Our measure of jitter is the standard deviation of latency.
\item[Versatility:] Is detection possible at ``difficult'' syllables?
\item[Ease of use:] How automated is the process of programming a detector?
\item[Cost:] What are the hardware and software requirements?
\end{description}

A variety of syllable-triggering systems have been used, but few have
been documented or characterised in detail.  In 1999,
detection was achieved by a group of IIR filters with hand-tuned logical
operators \cite{Leonardo1999}.  The system had a latency of 50 or 100 ms,
and accuracy and jitter were not reported.  As access to
computational resources has improved, approaches have changed: in
2009, hand-tuned filters were implemented on a Tucker-Davis Technologies digital signal processor, bringing latency down to around 4 ms \cite{Andalman2009}. But as with other filter-bank techniques, it
is not strictly a syllable detector but rather a pitch and timbre
detector---it cannot identify a frequency sweep, or distinguish a
short chirp from a long one---and thus requires careful selection of
target syllables.  Furthermore, the method is neither inexpensive nor,
based on our experience with a similar technique, accurate.
2009 saw the application of a neural network to a spectral image of
song \cite{Keller2009}.  They reported a jitter of 4.3 ms, but further implementation and
performance details are not available.  In 2011, stable portions of syllables were matched to spectral templates in 8-ms
segments \cite{Warren2011}.  This detector achieved a jitter of 4.5 ms, and false-negative and
false-positive rates of up to 2\% and 4\% respectively.  Hardware
requirements and ease of use were not reported.  In 2013, spectral
images of template syllables were compared to song using a correlation coefficient \cite{Skocik2013}.  With a
fast desktop (Intel i7 six-core) running Linux and equipped with a
National Instruments data acquisition card, it boasts a hardware-only
(not accounting for the time taken to compute a match with a syllable)
latency and jitter of just a few microseconds, and the detection
computation they use should not much increase that.  They reported
false-negative rates around 4\% and 7\% for zebra finches and
Bengalese finches respectively, measured on a small dataset.  In much
other work, a syllable detector is alluded to, but not described.

@@

\begin{enumerate}
\item Record and align a corpus of training songs.  The technique has
  been published in \cite{Poole2012}. As few as 200 songs can yield
  acceptable results, but here we standardise on 1000-song training
  sets.
\item Choose one or more instants in the song that should create a trigger event, and train a
  neural network to recognise them. This step is carried out offline.  While any neural network software would produce similar results, we used MATLAB 2015b's neural network toolbox.
\item Once trained and saved, the neural network is used by a realtime
  detection programme that listens to an audio signal and indicates detection of the target
  syllables via a TTL pulse.  We present three detection implementations, in MATLAB, LabVIEW, and Swift, that trade off hardware requirements, ease of maintenance, and performance.
\end{enumerate}
This method makes the following contributions:
\begin{itemize}
\item Fast: sub-millisecond latencies, with jitter around 2 ms.
\item Accurate: false negative rates under 1\% and false positive rates under 0.005\% for a variety of syllables.
\item State-of-the-art performance with default parameters.
\item Requires almost no programming experience.
\item Runs on inexpensive hardware.
\item Described in detail here, with reference implementations provided and benchmarked.
\end{itemize}

@@

We begin with a few hundred recordings of a given bird's song, as well
as calls, cage noise, and other non-song audio data.  Male zebra finch
song is most highly stereotyped when a female is present (``directed
song'') and slightly more variable otherwise (``undirected''); we
train and test on undirected song since this is both more commonly
studied and more challenging.  Recordings were made inside a
sound-isolating chamber in which was mounted a recording microphone
(Audio-Technica AT803), using methods similar to those described in
\cite{Tchernichovski2011SAP}, Chapter 2.  The songs were time-aligned
as described in \cite{Poole2012}.

@@

Six examples of chosen target moments in the song, with recognition
regions $F =$ 1--8 kHz and $T=30$ ms, are shown in \fig{fig:song}.

@@

With inputs well outside the space on which a neural network has been
trained, its outputs will be essentially random. In order to reduce
the false positive rate it is necessary to provide negative training
examples that include silence, cage noise, wing flapping, non-song
vocalisations, and perhaps songs from other birds.  Although it will
depend on the makeup of the non-song data, we have found that training
with as low as a 1:1 ratio of non-song to song---or roughly 10
minutes of non-song audio---yields excellent results on most birds.

@@

During runtime, the network may produce above-threshold responses to
nearby frames.  Thus, after the first response, subsequent responses
are suppressed for 100 ms.  However, for the accuracy measurements
presented here, we used the un-de-bounced network response.

@@

This implementation requires special software and hardware: LabVIEW
from National Instruments---we used 2014 service pack 1---and a data acquisition card---we use the
National Instruments PCI-6251 card on a PC with an Intel Core i5-4590
processor at 3.7GHz (a relatively low-end machine) with 24 gigabytes
of RAM, running Microsoft Windows 8.1 Pro and Windows 10.

@@

We define the accuracy of the network based on its classification
performance per frame. In order to avoid the apparent problem of every
non-detected non-matching frame counting as a true negative, we also
tried defining accuracy on a per-song basis, such that a song-length
sample without the target syllable counted as a single true negative.
Computing the optimal output thresholds on a per-frame rather than a
per-song basis resulted in higher thresholds and thus a slight
reduction in both the false-positive and true-positive rates
(reversible by increasing $C_n$), while also providing a valid
definition of false-positive rate for data streams that had not been
segmented into song-sized chunks and thus allowing easy interpretation
of false-positive rates from calls, cage noise, etc.

@@

\fig{fig:song} shows a typical song. The image is produced by
averaging the aligned spectrograms for the bird lny64.  Six target
trigger points are shown for this song, at 150, 200, 250, 300, 350,
and 400 milliseconds after the beginning of the aligned samples, as
indicated by the red lines. In discussion, we will refer to these six
as $t^*_1\ldots t^*_6$ respectively.  To avoid overlap in the graphic, the rectangles show
a shorter recognition window than our standard: here 30 ms, but still
spanning 1--8 kHz, for each target time in the song.

@@

\begin{table}
  % replot_accuracies_concatanated('tex', true) for confusion_log_perf.txt (1.5 ms).
  \begin{tabular}{lc|cc}
    Bird & Target time (ms) & \% True positives & \% False positives \\ \hline
lny64 &         150 $[t^*_1]$ &         99.66 &         0.0026 \\ % /Volumes/Data/song/lny64/confusion_log_perf.txt
      &         200 $[t^*_2]$ &         99.75 &         0.00052 \\ %                                               
      &         250 $[t^*_3]$ &         99.79 &         0.00051 \\ %                                               
      &         300 $[t^*_4]$ &         99.70 &         0.00077 \\ %                                               
      &         350 $[t^*_5]$ &         99.73 &         0.00022 \\ %                                               
      &         400 $[t^*_6]$ &         99.52 &         0.0010 \\ %                                                
lny42 &         380 &   99.09 &         0.0011 \\ % /Volumes/Data/song/lny42/confusion_log_perf.txt          
lny46 &         140 &   99.17 &         0.0037 \\ % /Volumes/Data/song/lny46/confusion_log_perf.txt          
lny4rb &        250 &   99.37 &         0.0026 \\ % /Volumes/Data/song/lny4rb/confusion_log_perf.txt        
lr12 &  180 &   99.52 &         0.0033 \\ % /Volumes/Data/song/lr12/all/confusion_log_perf.txt        
lr13 &  300 &   99.83 &         0.0013 \\ % /Volumes/Data/song/lr13/all/confusion_log_perf.txt        
lr28 &  140 &   99.73 &         0.0028 \\ % /Volumes/Data/song/lr28/all/confusion_log_perf.txt        
lr77 &  320 &   99.58 &         0.0028  % /Volumes/Data/song/lr77/all/confusion_log_perf.txt          
  \end{tabular}
  \caption{Mean values for the detection accuracies shown in \fig{fig:accuracies}.}
  \label{table:accuracies}
\end{table}

@@

Changing the FFT interval to 2 ms significantly reduces the
computational requirement during learning. For most syllables, results
were identical, but some syllables are slightly easier to detect and
some slightly more difficult, with no significant difference on
average.

@@

  \begin{description}
  \item[Installation:]\hfill
    \begin{itemize}
    \item Clone the Github repository onto your computer.
    \item You also need the following packages from MATLAB File Exchange somewhere in your MATLAB path:
      \begin{description}
      \item[Significant Figures] by Teck Por Lam.  File ID \#10669.
      \item[Generate maximally perceptually-distinct colors] by Tim Holy.  File ID \#29702.
      \end{description}
    \end{itemize}
  \item[Usage:]\hfill
    \begin{itemize}
    \item The main programme is {\tt learn\_detector.m}.
    \item The top section of {\tt learn\_detector.m}
      describes the configuration variables.
    \item The current directory should contain the data in {\tt
      song.mat}. The format for this file is specified in {\tt
      README.md}: a MATLAB file containing aligned song, nonsong
      (calls, cage noise, etc), and the audio sampling frequency.
    \item {\tt learn\_detector} also looks in the current directory for
      a configuration file {\tt params.m}, described in {\tt
        README.md}.  If the file is empty, the programme will display
      an image showing the average of the song spectrograms (frequency
      vs.~time), tell the user how to define syllables of interest,
      and exit.  If {\tt times\_of\_interest\_ms} is defined, the
      programme will train the neural network.  Optionally, {\tt
        params.m} can also override any of the configuration
      parameters at the top of {\tt learn\_detector.m}.
    \item Configuration may also be controlled by specifying
      parameter-value pairs to the function call in the common MATLAB
      fashion; for example, {\tt
        learn\_detector('times\_of\_interest\_ms', [150 200])}.
    \item The software will produce a neural network file, {\tt
      detector\_bird\_times\_\dots.mat} (the ellipses signify other
      information encoded in the filename).  It will also produce the
      audio test file described in Methods.
    \end{itemize}
  \end{description}
